Mixed effect tweedie model in TMB

### 2020-06-18 analysis/tmb-sensor-explore/2-fit-temperature.Rmd
rerun climate models with current sdmTMB coding

plan(multiprocess)
```

@@

```{r}
ggplot(dplyr::filter(d_trawl1, shallow==1), aes(SST_scaled,temperature_c, colour = -depth)) + geom_point()
```

@@

```{r eval = FALSE}
library(gbm)

@@

## Just compare fixed slope models with simplest random effect structure
```{r eval=F}
cv_args <- generate_cv_args(
  formula = list(
    temperature_c ~ 0 + SST_scaled*shallow + depth_scaled,
    temperature_c ~ 0 + SST_scaled*shallow + depth_scaled + depth_scaled2,
    temperature_c ~ 0 + SST_scaled*shallow + depth_scaled + depth_scaled2 + depth_scaled3,
    temperature_c ~ 1 + SST_scaled + depth_scaled,
    temperature_c ~ 1 + SST_scaled + depth_scaled + depth_scaled2,
    temperature_c ~ 1 + SST_scaled + depth_scaled + depth_scaled2 + depth_scaled3,
    temperature_c ~ 0 + as.factor(year) + depth_scaled,
    temperature_c ~ 0 + as.factor(year) + depth_scaled + depth_scaled2,
    temperature_c ~ 0 + as.factor(year) + depth_scaled + depth_scaled2 + depth_scaled3,
    temperature_c ~ 0 + as.factor(year) + SST_scaled + depth_scaled,
    temperature_c ~ 0 + as.factor(year) + SST_scaled + depth_scaled + depth_scaled2,
    temperature_c ~ 0 + as.factor(year) + SST_scaled + depth_scaled + depth_scaled2 + depth_scaled3,
    temperature_c ~ 0 + as.factor(year) + SST_scaled * depth_scaled,
    temperature_c ~ 0 + as.factor(year) + SST_scaled * depth_scaled + SST_scaled * depth_scaled2,
    temperature_c ~ 0 + as.factor(year) + SST_scaled * depth_scaled + SST_scaled * depth_scaled2 + SST_scaled * depth_scaled3
  ),
  time_varying = list(NULL),
  ar1_fields = c(FALSE),
  include_spatial = c(FALSE),
  n_knots = c(100),
  seed = c(1111, 2222, 3333, 4444)
)

@@

## Compare time-varying slope models still with simplest random effect structure
All of top models for any seed contained year and none showed a linear effect of depth
All are better than fixed slope models
```{r}
cv_args <- generate_cv_args(
  formula = list(
    #temperature_c ~ 0 + SST_scaled*shallow,
    # temperature_c ~ 0 + as.factor(year) + SST_scaled*shallow,
    # #temperature_c ~ 1 + SST_scaled,
    # temperature_c ~ 0 + as.factor(year) + SST_scaled,
    temperature_c ~ 0 + as.factor(year)
  ),
  time_varying = list(
    # ~ 0 + depth_scaled + depth_scaled2 + depth_scaled3,
    ~ 0 + depth_scaled + depth_scaled2,
  ),
  ar1_fields = c(TRUE),
  include_spatial = c(TRUE),
  n_knots = c(400), # keep same n_knots
  seed = c(1111, 2222, 3333, 4444) # keep same seeds
)

@@

## Now vary random effect configuation for top model(s) 
a separate spatial field is in all models with highest likelihoods for each seed
AR1 doesn't make much difference (differences between years appear independent)
```{r}
cv_args <- generate_cv_args(
  formula = list(
    temperature_c ~ 0 + as.factor(year) + SST_scaled*shallow,
    temperature_c ~ 0 + as.factor(year) + SST_scaled,
    temperature_c ~ 0 + as.factor(year)
  ),
  time_varying = list(
    ~ 0 + depth_scaled + depth_scaled2
  ),
  ar1_fields = c(FALSE, TRUE),
  include_spatial = c(FALSE, TRUE),
  n_knots = c(100),
  seed = c(1111, 2222, 3333, 4444)
)

@@

## Change knot number

```{r}
cv_args <- generate_cv_args(
    formula = list(
    # temperature_c ~ 0 + as.factor(year) + SST_scaled,
    temperature_c ~ 0 + as.factor(year)
  ),
  time_varying = list(
    ~ 0 + depth_scaled + depth_scaled2
  ),
  ar1_fields = c(FALSE,TRUE),
  include_spatial = c(TRUE),
  n_knots = c(140,180,220,250),
  #seed = c(1111, 2222, 3333, 4444) 
  seed = c(5555, 6666, 7777, 8888, 9999)
)

@@

## Change seeds 
Also increased knot number too 
Models without SST_scaled are usually better

```{r}
cv_args <- generate_cv_args(
    formula = list(
    temperature_c ~ 0 + as.factor(year) + SST_scaled*shallow,
    temperature_c ~ 0 + as.factor(year) + SST_scaled,
    temperature_c ~ 0 + as.factor(year)
  ),
  time_varying = list(
    ~ 0 + depth_scaled + depth_scaled2
  ),
  ar1_fields = c(FALSE, TRUE),
  include_spatial = c(TRUE, FALSE),
  n_knots = c(130),
  seed = c(5555, 6666, 7777, 8888, 9999)
)

@@

Current top model

```{r}
spde <- make_spde(d_trawl$X, d_trawl$Y, n_knots = 400)
plot_spde(spde)
```

```{r}
m_temp_rw <- sdmTMB(d_trawl,
  temperature_c ~ 0 + as.factor(year), #+ DOY_scaled + as.factor(ssid)
  time_varying = ~ 0 + depth_scaled + depth_scaled2,
  ar1_fields = TRUE, # changed AR1_field to true
  include_spatial = TRUE, 
  time = "year", spde = spde, family = gaussian(link = "identity"),
  silent = FALSE
)

saveRDS(m_temp_rw, file = (here::here("analysis/tmb-sensor-explore/models/model-temp-all-years-400kn.rds")))

@@

```{r}
nd_all <- readRDS(paste0("../VOCC/data/nd_all_synoptic.rds")) %>% filter(year<2019)
unique(nd_all$year)
glimpse(nd_all)
predict_temp <- predict(m_temp_rw, newdata = nd_all)
saveRDS(predict_temp, file = "../VOCC/data/predicted_temp_allyears_revised")
p_nd <- predict_temp
#p_nd <- readRDS("../VOCC/data/predicted_temp_allyears.rds")
# View(p_nd)
```

```{r}
p_nd <- readRDS("../VOCC/data/predicted_temp_allyears_revised")
# p_nd <- readRDS("~/github/dfo/gfranges/analysis/VOCC/data/predicted_temp_allyears_revised.rds")

@@

pred_1 <- plot_map_raster(p_nd1, "est") +
  scale_fill_viridis_c(labels = labels, 
    #trans = "sqrt",
    option = "C",
    #breaks = c(0,2,4,6,8), 
    limits = c(5, 11), 
    na.value = "red") +
  facet_wrap(~year) +
  labs(fill = "°C") + 
  theme(legend.position = "none") +
  ggtitle("Prediction (fixed effects + all random effects)") 
pred_1
```

>>>>

# plan(multiprocess)
```

@@

```{r}
ggplot(dplyr::filter(d_trawl, shallow==1), aes(SST_scaled,temperature_c, colour = -depth)) + geom_point()
```

@@

```{r, eval = FALSE}
library(gbm)

@@

## Just compare fixed slope models with simplest random effect structure
```{r, eval=F}
cv_args <- generate_cv_args(
  formula = list(
    temperature_c ~ 0 + SST_scaled*shallow + depth_scaled,
    temperature_c ~ 0 + SST_scaled*shallow + depth_scaled + depth_scaled2,
    temperature_c ~ 0 + SST_scaled*shallow + depth_scaled + depth_scaled2 + depth_scaled3,
    temperature_c ~ 1 + SST_scaled + depth_scaled,
    temperature_c ~ 1 + SST_scaled + depth_scaled + depth_scaled2,
    temperature_c ~ 1 + SST_scaled + depth_scaled + depth_scaled2 + depth_scaled3,
    temperature_c ~ 0 + as.factor(year) + depth_scaled,
    temperature_c ~ 0 + as.factor(year) + depth_scaled + depth_scaled2,
    temperature_c ~ 0 + as.factor(year) + depth_scaled + depth_scaled2 + depth_scaled3,
    temperature_c ~ 0 + as.factor(year) + SST_scaled + depth_scaled,
    temperature_c ~ 0 + as.factor(year) + SST_scaled + depth_scaled + depth_scaled2,
    temperature_c ~ 0 + as.factor(year) + SST_scaled + depth_scaled + depth_scaled2 + depth_scaled3,
    temperature_c ~ 0 + as.factor(year) + SST_scaled * depth_scaled,
    temperature_c ~ 0 + as.factor(year) + SST_scaled * depth_scaled + SST_scaled * depth_scaled2,
    temperature_c ~ 0 + as.factor(year) + SST_scaled * depth_scaled + SST_scaled * depth_scaled2 + SST_scaled * depth_scaled3
  ),
  time_varying = list(NULL),
  ar1_fields = c(FALSE),
  include_spatial = c(FALSE),
  n_knots = c(100),
  seed = c(1111, 2222, 3333, 4444)
)

@@

## Compare time-varying slope models still with simplest random effect structure
All of top models for any seed contained year and none showed a linear effect of depth
All are better than fixed slope models
```{r, eval=FALSE}
cv_args <- generate_cv_args(
  formula = list(
    #temperature_c ~ 0 + SST_scaled*shallow,
    # temperature_c ~ 0 + as.factor(year) + SST_scaled*shallow,
    # #temperature_c ~ 1 + SST_scaled,
    # temperature_c ~ 0 + as.factor(year) + SST_scaled,
    temperature_c ~ 0 + as.factor(year)
  ),
  time_varying = list(
    # ~ 0 + depth_scaled + depth_scaled2 + depth_scaled3,
    ~ 0 + depth_scaled + depth_scaled2,
  ),
  ar1_fields = c(TRUE),
  include_spatial = c(TRUE),
  n_knots = c(400), # keep same n_knots
  seed = c(1111, 2222, 3333, 4444) # keep same seeds
)

@@

## Now vary random effect configuation for top model(s) 
a separate spatial field is in all models with highest likelihoods for each seed
AR1 doesn't make much difference (differences between years appear independent)
```{r, eval=FALSE}
cv_args <- generate_cv_args(
  formula = list(
    temperature_c ~ 0 + as.factor(year) + SST_scaled*shallow,
    temperature_c ~ 0 + as.factor(year) + SST_scaled,
    temperature_c ~ 0 + as.factor(year)
  ),
  time_varying = list(
    ~ 0 + depth_scaled + depth_scaled2
  ),
  ar1_fields = c(FALSE, TRUE),
  include_spatial = c(FALSE, TRUE),
  n_knots = c(100),
  seed = c(1111, 2222, 3333, 4444)
)

@@

## Change knot number
```{r, eval=FALSE}
cv_args <- generate_cv_args(
    formula = list(
    # temperature_c ~ 0 + as.factor(year) + SST_scaled,
    temperature_c ~ 0 + as.factor(year)
  ),
  time_varying = list(
    ~ 0 + depth_scaled + depth_scaled2
  ),
  ar1_fields = c(FALSE,TRUE),
  include_spatial = c(TRUE),
  n_knots = c(140,180,220,250),
  #seed = c(1111, 2222, 3333, 4444) 
  seed = c(5555, 6666, 7777, 8888, 9999)
)

@@

## Change seeds 
Also increased knot number too 
Models without SST_scaled are usually better
```{r, eval=FALSE}
cv_args <- generate_cv_args(
    formula = list(
    temperature_c ~ 0 + as.factor(year) + SST_scaled*shallow,
    temperature_c ~ 0 + as.factor(year) + SST_scaled,
    temperature_c ~ 0 + as.factor(year)
  ),
  time_varying = list(
    ~ 0 + depth_scaled + depth_scaled2
  ),
  ar1_fields = c(FALSE, TRUE),
  include_spatial = c(TRUE, FALSE),
  n_knots = c(130),
  seed = c(5555, 6666, 7777, 8888, 9999)
)

@@

Current top model
```{r}
spde <- make_spde(d_trawl$X, d_trawl$Y, n_knots = 500)
plot_spde(spde)
```

```{r}
m_temp_rw <- sdmTMB(
  temperature_c ~ 0 + as.factor(year), #+ DOY_scaled + as.factor(ssid)
  time_varying = ~ 0 + depth_scaled + depth_scaled2,
  data = d_trawl,
  ar1_fields = TRUE, # changed AR1_field to true
  include_spatial = TRUE, 
  time = "year", spde = spde, family = gaussian(link = "identity"),
  silent = FALSE
)

saveRDS(m_temp_rw, file = (here::here("analysis/tmb-sensor-explore/models/model-temp-all-years-500kn.rds")))

@@

```{r}
nd_all <- readRDS(paste0("../VOCC/data/nd_all_synoptic.rds")) %>% filter(year<2019)
unique(nd_all$year)
glimpse(nd_all)
predict_temp <- predict(m_temp_rw, newdata = nd_all)
saveRDS(predict_temp, file = "../VOCC/data/predicted_temp_allyears_revised2")
p_nd <- predict_temp
#p_nd <- readRDS("../VOCC/data/predicted_temp_allyears.rds")
# View(p_nd)
```

```{r}
p_nd <- readRDS("../VOCC/data/predicted_temp_allyears_revised.rds")
# p_nd <- readRDS("~/github/dfo/gfranges/analysis/VOCC/data/predicted_temp_allyears_revised.rds")
p_nd2 <- readRDS("../VOCC/data/predicted_temp_allyears_revised2")

plot(p_nd$est ~ p_nd2$est)

@@

pred_1 <- plot_map_raster(p_nd1, "est") +
  scale_fill_viridis_c(labels = labels, 
    #trans = "sqrt",
    option = "C",
    #breaks = c(0,2,4,6,8), 
    limits = c(4, 11),
    na.value = "red") +
  facet_wrap(~year) +
  labs(fill = "°C") + 
  # theme(legend.position = "none") +
  ggtitle("Prediction (fixed effects + all random effects)") 
pred_1
```

### 2020-06-18 analysis/tmb-sensor-explore/3-fit-do.Rmd
rerun climate models with current sdmTMB coding

d_trawl1 <- d_trawl %>% filter(!(year == 2016 & ssid ==4)) #%>% filter(ssid != 16) %>% filter(ssid != 4)    
#test <- d_trawl1 %>% filter(year == 2016 & ssid ==16) 
spde <- make_spde(d_trawl1$X, d_trawl1$Y, n_knots = 800)
plot_spde(spde)
```

@@

Add temperature predictions to newdata grid for DO predictions
```{r}
temperature <- readRDS(here::here("analysis/VOCC/data/predicted_temp_allyears.rds"))
years <- unique(d_trawl1$year) 
temperature <- filter(temperature, year %in% years)
unique(temperature$year)

@@

    do_model_all_DOY <- sdmTMB::sdmTMB(d_trawl1,
      log_do ~ 0 + as.factor(year) + temp_scaled + temp_scaled2 + DOY_scaled,
      time_varying = ~ 0 + depth_scaled + depth_scaled2 + depth_scaled3,
      time = "year", spde = spde,
      family = gaussian(link = "identity"),
      ar1_fields = TRUE,
      include_spatial = TRUE,
      enable_priors = FALSE,
      control = sdmTMBcontrol(step.min = 0.01, step.max = 1),
      silent = FALSE
    )
    endtime1 <- Sys.time()
    time1 <- round(starttime1 - endtime1)

@@

Other models tried:

@@

Predict DO for whole grid using quadratic of predicted temperature values and mean DOY
```{r eval=FALSE}
p_nd <- predict(do_model_DOY, newdata = newdata)
p_nd$do_est <- exp(p_nd$est)

@@

model1 <- readRDS("~/github/dfo/gfranges/analysis/tmb-sensor-explore/models/model-temp-2008-on-used.rds")
#model2 <- readRDS("~/github/dfo/gfranges/analysis/tmb-sensor-explore/models/model-do-all-DOY-used.rds")
#update model to exclude WCVI 2016 outliers
model2 <- readRDS("~/github/dfo/gfranges/analysis/tmb-sensor-explore/models/model-do-without-wcvi2016.rds")

nd_all <- readRDS(paste0("data/nd_all_synoptic.rds"))
nd <- filter(nd_all, year >2007) %>% filter(year<2019)
pred_temp <- predict(model1, newdata = nd)
predtemp <- pred_temp %>% mutate(temp_scaled = est, temp_scaled2 = est^2, depth_scaled3 = depth_scaled^3) %>% select(-est, -est_non_rf, -est_rf, -omega_s, -zeta_s, -epsilon_st)
predtemp$DOY_scaled <- 0
pred_do <- predict(model2, newdata = predtemp)
pred_do$temp <- pred_do$temp_scaled*1.296 + 6.49

@@

saveRDS(pred_do, file="~/github/dfo/gfranges/analysis/VOCC/data/predicted-DO-new.rds")

@@

```{r}
#p_nd <- readRDS((here::here("analysis/tmb-sensor-explore/models/predicted-DO.rds")))
p_nd <- readRDS((here::here("analysis/VOCC/data/predicted-DO-new.rds")))

@@

pred_1 <- plot_map_raster(p_nd1, "do_est") +
  scale_fill_viridis_c(labels = labels, breaks = c(0,2,4,6,8), limits = c(0, 9), na.value = "red") +
  facet_wrap(~year) +
  labs(fill = "ml/L DO") +
  ggtitle("Prediction (fixed effects + all random effects)") + 
  theme(legend.position = c(.85,.25))
pred_1
```

```{r}
p_nd4 <- p_nd %>% filter(ssid==4) #%>% filter(year!=2016)
pred_4 <- plot_map_raster(p_nd4, "do_est") +
  scale_fill_viridis_c(labels = labels, breaks = c(0,2,4,6,8), limits = c(0, 9), na.value = "red")+
  facet_wrap(~year) +
  labs(fill = "ml/L DO") +
  #ggtitle(" ")+ 
  theme(legend.position = "none")
pred_4
```

>>>>

d_trawl1 <- d_trawl %>% filter(!(year == 2016 & ssid ==4)) #%>% filter(ssid != 16) %>% filter(ssid != 4)    
#test <- d_trawl1 %>% filter(year == 2016 & ssid ==16) 
spde <- make_spde(d_trawl1$X, d_trawl1$Y, n_knots = 800) ### TODO: rerun with 500
plot_spde(spde)
```

@@

Add temperature predictions to newdata grid for DO predictions
```{r eval=FALSE}
temperature <- readRDS(here::here("analysis/VOCC/data/predicted_temp_allyears_revised2"))
years <- unique(d_trawl1$year) 
temperature <- filter(temperature, year %in% years)
unique(temperature$year)

@@

    do_model_all_DOY <- sdmTMB::sdmTMB(
      log_do ~ 0 + as.factor(year) + temp_scaled + temp_scaled2 + DOY_scaled,
      data = d_trawl1,
      time_varying = ~ 0 + depth_scaled + depth_scaled2 + depth_scaled3,
      time = "year", 
      spde = spde,
      family = gaussian(link = "identity"),
      ar1_fields = TRUE,
      include_spatial = TRUE,
      enable_priors = FALSE,
      control = sdmTMBcontrol(step.min = 0.01, step.max = 1),
      silent = FALSE
    )
    endtime1 <- Sys.time()
    time1 <- round(starttime1 - endtime1)

@@

Other models tried:
```{r eval=FALSE}
do_model_DOY <- readRDS(here::here("analysis/tmb-sensor-explore/models/model-do-without-wcvi2016.rds"))
do_model_DOY
#p_nd <- predict(do_model_DOY, newdata = newdata)
get_diag(do_model_DOY, response = "log_do", variable = "depth_scaled")
```

@@

Predict DO for whole grid using quadratic of predicted temperature values and mean DOY
```{r eval=FALSE}
do_model_all <- readRDS(here::here("analysis/tmb-sensor-explore/models/model-do-without-wcvi2016-poly.rds"))
p_nd <- predict(do_model_all_DOY, newdata = newdata)
p_nd$do_est <- exp(p_nd$est)

@@

# model1 <- readRDS("~/github/dfo/gfranges/analysis/tmb-sensor-explore/models/model-temp-2008-on-used.rds")
#model2 <- readRDS("~/github/dfo/gfranges/analysis/tmb-sensor-explore/models/model-do-all-DOY-used.rds")

# fix RW additon error
model1 <- readRDS(here::here("analysis/tmb-sensor-explore/models/model-temp-all-years-500kn.rds"))
# update model to exclude WCVI 2016 outliers
model2 <- readRDS(here::here("analysis/tmb-sensor-explore/models/model-do-without-wcvi2016.rds"))

nd_all <- readRDS(here::here("analysis/VOCC/data/nd_all_synoptic.rds")) %>% filter(year<2019)

pred_temp <- predict(model1, newdata = nd_all)

# get scaling used in model
d <- all_sensor %>%
  # filter missing do data 
  dplyr::filter(!is.na(do_mlpl)) %>%
  # remove 2007 data because clearly faulty or scaled differently
  dplyr::filter(year>2007)
d$temp_scaled <- scale(d$temperature_c)
attributes(d$temp_scaled)

predtemp <- pred_temp %>% mutate(temp_scaled = (est - 6.49 )/1.296, temp_scaled2 = ((est - 6.49 )/1.296)^2, depth_scaled3 = depth_scaled^3) %>% select(-est, -est_non_rf, -est_rf, -omega_s, -zeta_s, -epsilon_st)
predtemp <- predtemp %>% filter(year > 2007) 
predtemp$DOY_scaled <- 0

pred_do <- predict(model2, newdata = predtemp)

pred_do$temp <- pred_do$temp_scaled*1.296 + 6.49

@@

saveRDS(pred_do, file=here::here("analysis/VOCC/data/predicted-DO-june-2020.rds"))

@@

```{r}
#p_nd <- readRDS((here::here("analysis/tmb-sensor-explore/models/predicted-DO.rds")))
# p_nd <- readRDS((here::here("analysis/VOCC/data/predicted-DO-new.rds")))

p_nd <- readRDS((here::here("analysis/VOCC/data/predicted-DO-june-2020.rds")))

@@

pred_1 <- plot_map_raster(p_nd1, "do_est") +
  scale_fill_viridis_c(labels = labels, breaks = c(0,2,4,6,8), 
    limits = c(1, 8), 
    na.value = "red") +
  facet_wrap(~year) +
  labs(fill = "ml/L DO") +
  ggtitle("Prediction (fixed effects + all random effects)") + 
  theme(legend.position = c(.85,.25))
pred_1
```

```{r}
p_nd4 <- p_nd %>% filter(ssid==4) #%>% filter(year!=2016)
pred_4 <- plot_map_raster(p_nd4, "do_est") +
  scale_fill_viridis_c(labels = labels, breaks = c(0,2,4,6,8), limits = c(1, 7), 
     na.value = "red")+
  facet_wrap(~year) +
  labs(fill = "ml/L DO") +
  #ggtitle(" ")+ 
  theme(legend.position = "none")
pred_4
```
